---
title: "29May2025"
author: "Nate Talampas"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The dataset AirPassengers includes monthly totals of international flights between 1949-1960. Fit a regression model to the data.

```{r}
data(AirPassengers) # Calling the built-in R data
head(AirPassengers)
```

# Plot of the Data (log transformation)

Let $Y_t$ show the total number of international passengers at time $t$.

```{r}
plot(AirPassengers)
```

```{r}
plot(log(AirPassengers))
```

# Models with either trend or seasonal component

$log(Y_t) = \beta_0 + \beta_1 t + R_t$


```{r}
tim <- time(AirPassengers) #creates vector of time that TS was collected
cycle(AirPassengers) #gives the positions in the cycle of each observation
month <- as.factor(cycle(AirPassengers))
reg1 <- lm(log(AirPassengers)~tim)

plot(log(AirPassengers))
points(tim,predict.lm(reg1),type='l',col='red')
```

$log(Y_t) = \beta_0 + \beta_1 x_1 + ... \beta_{11} x_{11} + R_t$

```{r}
tim <- time(AirPassengers) #creates vector of time that TS was collected
cycle(AirPassengers) #gives the positions in the cycle of each observation
month <- as.factor(cycle(AirPassengers))
reg1 <- lm(log(AirPassengers)~ month)

plot(log(AirPassengers))
points(tim,predict.lm(reg1),type='l',col='red')
```

# Regression model

- Use a model with both trend and seasonal component
$log(Y_t) = m_t + S_t + R_t$

- We fit the model

$$ \log(Y_t) = \beta_0 + \beta_1 t + \mathrm{seasonal\ component} + R_t $$

where $R_t$ is the noise $R_t ~ N(0, \sigma^2) (independent)$

- Having monthly data, the seasonal component includes 11 binary explanatory variables (why?)

I think because month is a factor variable and needs a reference.

# Linear Trend

$log(Y_t) = \beta_0 + \beta_1 t + \text{seasonal component} + R_t$

$\text{seasonal component} = \beta_2 X_1 + \beta_3 X_2 + ... + \beta_{12} X_{11}$

where $X_1, X_2, ..., X_{11}$ are binary variables. i.e. $X_i \in \{0, 1\}$

To introduce a seasonal component with period, $d$, $d-1$ parameters are added to the regression model

# The Fit of the Model

$$ Y_t = \text{linear trend} + \text{seasonal component} + R_t$$

```{r}
#tim <- time(AirPassengers) #creates vector of time that TS was collected
#cycle(AirPassengers) #gives the positions in the cycle of each observation
month <- as.factor(cycle(AirPassengers))
reg1 <- lm(log(AirPassengers)~tim+month)

plot(log(AirPassengers))
points(tim,predict.lm(reg1),type='l',col='red')
```


```{r}
par(mfrow = c(2, 2))  # 2x2 layout

# 1. Fitted vs Residuals
plot(reg1$fitted.values, reg1$residuals,
     main = "Fitted vs Residuals",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# 2. Normal Q-Q plot
qqnorm(reg1$residuals, main = "Normal Q-Q Plot")
qqline(reg1$residuals, col = "red")

# 3. Index vs Residuals
plot(seq_along(reg1$residuals), reg1$residuals,
     main = "Index vs Residuals",
     xlab = "Index", ylab = "Residuals", type = "p")
abline(h = 0, col = "red", lty = 2)

# 4. Residuals Time Series
acf(reg1$residuals, main = "ACF of Residuals")
```

```{r}
par(mfrow = c(2, 2))  # 2x2 layout
plot(reg1$fitted, reg1$residuals) # plot of fitted values vs residuals 
qqnorm(reg1$residuals) #qq-plot of residuals
qqline(reg1$residuals) # plotting the line, along which the dots in qq-plot should lie 
plot(reg1$residuals) # plotting the residuals vs time
abline(h=0,lty=2) # plotting a horizontal line at 0
acf(reg1$residuals) #sample acf plot of residuals

```

The residuals indicate the presence of a quadratic relationship.

Exponentiate the previous graph:

```{r}
plot(AirPassengers)
points(tim,exp(predict.lm(reg1)),type='l',col='red')
```

We don't want a consistent gap => correlation still happening.

# Quadratic Trend

$$ log(Y_t) = \beta_0 + \beta_1 t + \beta_2 t^2 + \text{seasonal component} + R_t$$
$$ \text{seasonal component} = \beta_3 X_1 + \beta_4 X_2 + ... + \beta_{13} X_{11}$$
```{r}
tim2 <- tim^2
reg2 <- lm(log(AirPassengers)~tim+tim2+month)
summary(reg2)

```

```{r}
plot(log(AirPassengers))
points(tim,predict.lm(reg2),type='l',col='red')
#dev.off()
```

Exponentiate the previous graph.

```{r}
plot(AirPassengers)
points(tim,exp(predict.lm(reg2)),type='l',col='red')
```

```{r}
par(mfrow = c(2, 2))  # 2x2 layout

# 1. Fitted vs Residuals
plot(reg2$fitted.values, reg2$residuals,
     main = "Fitted vs Residuals",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# 2. Normal Q-Q plot
qqnorm(reg2$residuals, main = "Normal Q-Q Plot")
qqline(reg2$residuals, col = "red")

# 3. Index vs Residuals
plot(seq_along(reg2$residuals), reg2$residuals,
     main = "Index vs Residuals",
     xlab = "Index", ylab = "Residuals", type = "p")
abline(h = 0, col = "red", lty = 2)

# 4. Residuals Time Series
acf(reg2$residuals, main = "ACF of Residuals")
```

